A Toolbox for Manipulating and Assessing Color Palettes for Statistical Graphics

### 2019-02-26 pkg/colorspace/paper/Toolbox/article.Rmd
first draft of paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@457 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

## Usage with ggplot2

@@

Subsequently, the same types of assessment are carried out for the sequential `"Purples 3"` palette
as employed above.

@@

```{r hcl-palettes-qualitative, fig.width = 5.5, fig.height = 5.0, out.width = "0.4\\textwidth", fig.cap = "Prespecified qualitative HCL palettes available in \\code{qualitative\\_hcl()} in \\pkg{colorspace}."}
hcl_palettes("qualitative", plot = TRUE)
```

@@

```{r hcl-palettes-sequentials, fig.width = 6.5, fig.height = 6.0, out.width = "0.5\\textwidth", fig.cap = "Prespecified sequential single-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
hcl_palettes("sequential (single-hue)", n = 7, plot = TRUE)
```

@@

```{r hcl-palettes-diverging, fig.width = 6.5, fig.height = 7.5, out.width = "0.7\\textwidth", fig.cap = "Prespecified diverging HCL palettes available in \\code{diverging\\_hcl()} in \\pkg{colorspace}."}
hcl_palettes("diverging", n = 7, plot = TRUE)
```

@@

```{r hcl-trajectories, echo = FALSE, results = "hide", fig.width = 7, fig.height = 2.5, fig.cap = "Foo."}
mylab <- function(x, y, text, ratio = 1, col = "black", ...) {
  rect(x - 0.045/ratio, y - 3.5, x + 0.04/ratio, y + 4, border = col, ...)
  text(x, y, text, col = col, ...)
}

@@

c2 <- 10
plot(1 - i, c2 - (c2 - c1) * i, xlim = c(0, 1), ylim = c(0, 100),
  xlab = "Intensity (i)", ylab = "Coordinate", main = "Linear",
  xaxs = "i", yaxs = "i", type = "l", lwd = 2, axes = FALSE)
axis(1, at = c(0, 0.5, 1), labels = c("1", "0.5", "0"))
axis(2)
box()
lines(1 - i, c2 - (c2 - c1) * i^1.6, lty = 2, lwd = 2)
mylab(0.06, c1 + 5, "c1")
mylab(0.94, c2 - 5, "c2")
mylab(0.68, 40, "p1", col = gray(0.35))
text(0.78, 40, "= 1", col = gray(0.35))
mylab(0.18, 40, "p1", col = gray(0.35))
text(0.30, 40, "= 1.6", col = gray(0.35))

@@

plot(1 - i, ci1, xlim = c(0, 1), ylim = c(0, 100),
  xlab = "Intensity (i)", ylab = "Coordinate", main = "Triangular",
  xaxs = "i", yaxs = "i", type = "l", lwd = 2, axes = FALSE)
axis(1, at = c(0, 0.5, 1), labels = c("1", "0.5", "0"))
axis(2)
box()
lines(1 - i, ci2, lwd = 2, lty = 2)
mylab(0.06, c1 - 5, "c1")
mylab(1 - maxat + 0.1, cmax + 4, "cmax", ratio = 0.5)
mylab(0.94, c2 - 5, "c2")
mylab(0.77, 40, "p1", col = gray(0.35))
text(0.87, 40, "= 1", col = gray(0.35))
mylab(0.27, 40, "p1", col = gray(0.35))
text(0.39, 40, "= 1.6", col = gray(0.35))
```

@@

The `hcl_palettes()` already come with a wide range of pre-defined palettes to which
customizations can be easily added. However, it might also be convenient to register
a custom palette so that it can subsequently be re-used with a new dedicated name.
This is supported by adding a `register = "..."` argument once to a call to
`qualitative_hcl()`, `sequential_hcl()`, or `diverging_hcl()`:

@@

And can be used subsequently in `qualitative_hcl()` as well as the
qualitative [ggplot2 color scales](ggplot2_color_scales.html), e.g.,

@@

* The number of colors in the palette that was used during registration is not
  actually stored and can be modified subsequently. The same holds for arguments
  `alpha` and `rev`.
* When registering a new palette with an old name that was already available
  previously, the old palette gets overwritten. We recommend not to overwrite
  the palettes that are pre-defined in the package (albeit it is technically possible).
* The registration of a palette is only stored for the current session. When
  \proglang{R} is restarted and/or the `colorspace` package reloaded, only the pre-defined
  palettes from the package are available. Thus, to make a palette permanently
  available a registration \proglang{R} code like
  `colorspace::qualitative_hcl(3, palette = "set2", l = 80, register = "myset")`
  can be placed in your `.Rprofile` or similar startup scripts.

@@

```{r divergingx-palettes, fig.width = 6.5, fig.height = 8.0, out.width = "0.7\\textwidth", fig.cap = "Foo."}
divergingx_palettes(n = 7, plot = TRUE)
```

@@

- \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003] as provided by \proglang{R} package \pkg{RColorBrewer} [@color:RColorBrewer].
- \pkg{CARTO} colors [@color:CARTO] as provided by \proglang{R} package \pkg{rcartocolor} [@color:rcartocolor].
- The viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
  as provided by \proglang{R} package \pkg{viridis} [@color:viridis].
- The scientific color maps of @color:Crameri:2018 as provided by \proglang{R} package \pkg{scico} [@color:scico].

See the discussion of [HCL-based palettes](hcl_palettes.html) for more details. In the
following `specplot()` is used to compare the [HCL spectrum](palette_visualization.html#hcl-and-rgb-spectrum)
of the original palettes (top swatch, solid lines) and their HCL-based approximations (bottom swatch, dashed lines).

```{r brewer-carto-viridis, results = "hide", echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "Foo."}
specplot(
  rev(RColorBrewer::brewer.pal(7, "YlGnBu")),
  sequential_hcl(7, "YlGnBu"),
  main = "ColorBrewer.org: YlGnBu"
)
specplot(
  viridis::viridis(7),
  sequential_hcl(7, "Viridis"),
  main = "viridis: Viridis"
)
specplot(
  rcartocolor::carto_pal(7, "ag_Sunset"),
  sequential_hcl(7, "ag_Sunset"),
  main = "CARTO: ag_Sunset"
)
specplot(
  viridis::plasma(7),
  sequential_hcl(7, "Plasma"),
  main = "viridis: Plasma"
)
```

@@

## Color swatches

The function `swatchplot()` is a convenience function for displaying
collections of palettes that can be specified as lists or matrices of
hex color codes. Essentially, it is just a call to the base graphics
`rect()` function but with heuristics for choosing default labels, margins,
spacings, borders, etc. These are selected to work well for `hcl_palettes()`
and might need further tweaking in future versions.

As a first simple example:
`swatchplot("Label" = sequential_hcl(5))`.

As a somewhat more complex example a `swatchplot()` is created with three matrices
of sequential color palettes of blues, purples, reds, and greens. For all palettes,
luminance increases monotonically to yield a proper sequential palette. However,
the hue and chroma handling is somewhat different to emphasize different parts of the
palette.

- Single-hue: In each palette the hue is fixed and chroma decreases monotonically
  (along with increasing luminance). This is typically sufficient to clearly bring
  out the extreme colors (dark/colorful vs.\ light gray).
- Single-hue (advanced): The hue is fixed (as above) but the chroma trajectory is
  triangular. Compared to the basic single-hue palette above this allows to better
  distinguish the colors in the middle and not only the extremes.
- Multi-hue (advanced): As in the advanced single-hue palette the chroma trajectory
  is triangular but additionally the hue varies slightly. This can further enhance
  the distinction of colors in the middle of the palette.

```{r swatch-brpg, fig.width = 11.5, fig.height = 3.5, fig.cap = "Foo."}
bprg <- c("Blues", "Purples", "Reds", "Greens")
swatchplot(
  "Single-hue"            = t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
  "Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
  "Multi-hue (advanced)"  = t(sapply(bprg,           sequential_hcl, n = 7)),
  nrow = 5, line = 5
)
```

## HCL (and RGB) spectrum

@@

The function first transforms a given color palette to its HCL (`polarLUV()`)
coordinates. As the hues for low-chroma colors are not (or only poorly) identified,
by default a smoothing is applied to the hues. Also, to avoid jumps from 0 to 360 or vice
versa, the hue coordinates are shifted suitably.

By default, the resulting trajectories in the HCL spectrum are visualized by a simple line plot:

@@

FIXME: Point to previous plots

Finally, the diverging `"Green-Brown"` palette is depicted in the left panel of Figure\ \ref{fig:specplot}.
It simply combines a green and a brown/yellow sequential single-hue palette, both with triangular chroma trajectory.
Hue is constant in each "arm" of the palette and the chroma/luminance trajectories are balanced between both arms.
In the center the palette passes through a light gray (with zero chroma) as the neutral value.
By including the corresponding RGB spectrum in the top panel, it also becomes apparent that
choosing such well-balanced palettes through trajectories in RGB color space is not straightforward.
This balanced palette -- based on relatively simple HCL trajectories -- is contrasted
with a poorly-balanced palette -- based on simple linear RGB trajectories in the right panel of
Figure\ \ref{fig:specplot}. This depicts the RGB and HCL spectrum of the 
(in)famous RGB rainbow palette. [See @color:Hawkins+McNeall+Stephenson:2014 for a
plea why the RGB rainbow palette should be avoided in almost all scientific graphics.]

```{r specplot, fig.width = 5, fig.height = 7, out.width = "0.49\\textwidth", fig.cap = "Foo."}
specplot(diverging_hcl(100, "Green-Brown"), rgb = TRUE)
specplot(rainbow(100), rgb = TRUE)
```

The RGB spectrum shows that the trajectories are quite simple in RGB space but the
HCL spectrum shows that the result is very unbalanced with shifts and kinks in both
chroma and (more importantly) luminance. This is why this palette is not suitable
for encoding underlying data in statistical graphics. See also the related discussion
of [color vision deficiency](color_vision_deficiency.html).

## Trajectories in HCL space

@@

Therefore, the `hclplot()` is another visualization of the HCL coordinates associated
with a palette. It does so by collapsing over one of the coordinates (either the hue H
or the luminance L) and displays a heatmap of colors combining the remaining
two dimensions. The coordinates for the given color palette are highlighted
to bring out its trajectory. In case the hue is really fixed (as in single-hue
sequential palettes) or the luminance is really fixed (as in the qualitative
palettes), collapsing is straightforward. However, when the coordinate that is
collapsed over is actually not constant in the palette, a simple bivariate linear
model is used to capture how the collapsed coordinate varies along with the
two displayed coordinates.

@@

- `type = "qualitative"` if luminance is approximately constant.
- `type = "sequential"` if luminance is monotonic.
- `type = "diverging"` if luminance is diverging with two monotonic "arms" in the trajectory.

Thus, for qualitative palettes - where luminance and chroma are fixed - the varying
hue is displayed in a projection onto the hue-chroma plane at a given fixed luminance
(Figure\ \ref{fig:hcl-qualitative}):

```{r hcl-qualitative, fig.width = 4, fig.height = 4, fig.cap = "Foo.", dev = "png", dpi = 150, out.width = "0.5\\textwidth"}
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

@@

```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, fig.cap = "Foo.", dev = "png", dpi = 150}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = 260, c = 80, l = c(35, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95), power = c(0.8, 1.4)))
```

Note that for H = 260 it is possible to go to dark colors (= low luminance) with
high chroma while this is not possible to the same extent for H = 245. Hence,
chroma has to be decreased when proceeding to the dark low-luminance colors.

@@

```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, fig.cap = "Foo.", dev = "png", dpi = 150}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = c(260, 220), c = c(50, 75, 0), l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = c(260, 60), c = 60, l = c(40, 95), power = 1))
hclplot(diverging_hcl(7, h = c(260, 0), c = 80, l = c(35, 95), power = 1))
```

@@

## Demonstration of statistical graphics

To demonstrate how different kinds of color palettes work in different
kinds of statistical displays, `demoplot()` provides a simple convenience
interface to some base graphics with (mostly artificial) data sets.
As a first overview, all built-in demos are displayed with the same
sequential heat colors palette: `sequential_hcl(5, "Heat")`.

```{r demoplot-all, echo = FALSE, fig.width = 8.5, fig.height = 8.5, fig.cap = "Foo."}
par(mfrow = c(3, 3))
cl <- sequential_hcl(5, "Heat")
for (i in c("map", "heatmap", "scatter", "spine", "bar", "pie", "perspective", "mosaic", "lines")) {
  demoplot(cl, type = i)
  mtext(side = 3, text = i, line = 1.5)
}
```

All types of demos can, in principle, deal with arbitrarily many colors from any palette
but clearly the graphics differ in various respects such as:

@@

```{r demoplot-qualitative, fig.width = 8.5, fig.height = 3.0, fig.cap = "Foo."}
par(mfrow = c(1, 3))
demoplot(qualitative_hcl(4, "Pastel 1"), type = "pie")
demoplot(qualitative_hcl(4, "Set 2"),    type = "scatter")
demoplot(qualitative_hcl(4, "Dark 3"),   type = "lines")
```

@@

```{r demoplot-sequential, fig.width = 8.5, fig.height = 3.0, fig.cap = "Foo."}
par(mfrow = c(1, 3))
demoplot(sequential_hcl(99, "Purple-Blue"), type = "heatmap")
demoplot(sequential_hcl(99, "Reds"),        type = "perspective")
demoplot(sequential_hcl( 4, "Viridis"),      type = "spine")
```

@@

```{r demoplot-diverging, fig.width = 8.5, fig.height = 3.0, fig.cap = "Foo."}
par(mfrow = c(1, 3))
demoplot(diverging_hcl(99, "Tropic", power = 2.5), type = "map")
demoplot(diverging_hcl( 5, "Green-Orange"),        type = "mosaic")
demoplot(diverging_hcl( 5, "Blue-Red 2"),          type = "bar")
```

@@

```{r demoplot-dark, fig.width = 8.5, fig.height = 6.0, fig.cap = "Foo."}
par(mfrow = c(2, 3), bg = "black")
demoplot(sequential_hcl(9, "Oslo"), "heatmap")
demoplot(sequential_hcl(9, "Turku"), "heatmap")
demoplot(sequential_hcl(9, "Inferno", rev = TRUE), "heatmap")
demoplot(qualitative_hcl(9, "Set 2"), "lines")
demoplot(diverging_hcl(9, "Berlin"), "scatter")
demoplot(diverging_hcl(9, "Cyan-Magenta", l2 = 20), "lines")
```

@@

Using the physiologically-based model for simulating color vision deficiency (CVD)
of @color:Machado+Oliveira+Fernandes:2009 different kinds of limitations can be
emulated: deuteranope (green cone cells defective), protanope (red cone cells defective),
and tritanope (blue cone cells defective).
The workhorse function to do so is `simulate_cvd()` which can take any vector
of valid \proglang{R} colors and transform them according to a certain CVD transformation
matrix and transformation equation. The transformation matrices have been established
by @color:Machado+Oliveira+Fernandes:2009 and are provided in objects `protanomaly_cvd`,
`deutanomaly_cvd`, and `tritanomaly_cvd`.

The convenience interfaces `deutan()`, `protan()`, and `tritan()` are the high-level functions for
simulating the corresponding kind of color blindness with a given severity (calling
`simulate_cvd()` internally).

>>>>

## Usage with ggplot2 {#sec:ggplot2}

@@

Subsequently, the same types of assessment are carried out in Figure\ \ref{fig:allplots-sequential}
for the sequential `"Purples 3"` palette as employed above.

@@

```{r hcl-palettes-qualitative, fig.width = 8.0, fig.height = 2.5, out.width = "0.8\\textwidth", fig.cap = "Prespecified qualitative HCL palettes available in \\code{qualitative\\_hcl()} in \\pkg{colorspace}."}
hcl_palettes("qualitative", plot = TRUE, nrow = 5)
```

@@

```{r hcl-palettes-sequentials, fig.width = 8.0, fig.height = 3.0, out.width = "0.8\\textwidth", fig.cap = "Prespecified sequential single-hue HCL palettes available in \\code{sequential\\_hcl()} in \\pkg{colorspace}."}
hcl_palettes("sequential (single-hue)", n = 7, plot = TRUE, nrow = 6)
```

@@

```{r hcl-palettes-diverging, fig.width = 8.0, fig.height = 4.75, out.width = "0.8\\textwidth", fig.cap = "Prespecified diverging HCL palettes available in \\code{diverging\\_hcl()} in \\pkg{colorspace}."}
hcl_palettes("diverging", n = 7, plot = TRUE, nrow = 10)
```

@@

```{r hcl-trajectories, echo = FALSE, results = "hide", fig.width = 7, fig.height = 2.5, fig.cap = "Types of trajectories to construct HCL color palettes, exemplified for the chroma coordinates, see Equations~\\ref{eq:constant}--\\ref{eq:triangular}."}
mylab <- function(x, y, text, ratio = 1, col = "black", ...) {
  rect(x - 0.045/ratio, y - 3.5, x + 0.04/ratio, y + 4, border = col, ...)
  text(x, y, text, col = col, ...)
}

@@

c2 <- 10
plot(1 - i, c2 - (c2 - c1) * i, xlim = c(0, 1), ylim = c(0, 100),
  xlab = "Intensity (i)", ylab = "Coordinate", main = "Linear",
  xaxs = "i", yaxs = "i", type = "l", lwd = 2, axes = FALSE)
axis(1, at = c(0, 0.5, 1), labels = c("1", "0.5", "0"))
axis(2)
box()
lines(1 - i, c2 - (c2 - c1) * i^1.6, lwd = 2, col = gray(0.5))
mylab(0.06, c1 + 5, "c1")
mylab(0.94, c2 - 5, "c2")
mylab(0.68, 40, "p1", col = gray(0.5))
text(0.78, 40, "= 1", col = gray(0.5))
mylab(0.18, 40, "p1", col = gray(0.5))
text(0.30, 40, "= 1.6", col = gray(0.5))

@@

plot(1 - i, ci1, xlim = c(0, 1), ylim = c(0, 100),
  xlab = "Intensity (i)", ylab = "Coordinate", main = "Triangular",
  xaxs = "i", yaxs = "i", type = "l", lwd = 2, axes = FALSE)
axis(1, at = c(0, 0.5, 1), labels = c("1", "0.5", "0"))
axis(2)
box()
lines(1 - i, ci2, lwd = 2, col = gray(0.5))
mylab(0.06, c1 - 5, "c1")
mylab(1 - maxat + 0.1, cmax + 4, "cmax", ratio = 0.5)
mylab(0.94, c2 - 5, "c2")
mylab(0.77, 40, "p1", col = gray(0.5))
text(0.87, 40, "= 1", col = gray(0.5))
mylab(0.27, 40, "p1", col = gray(0.5))
text(0.39, 40, "= 1.6", col = gray(0.5))
```

@@

The `hcl_palettes()` already come with a wide range of predefined palettes to which
customizations can be easily added. However, it might also be convenient to register
a custom palette so that it can subsequently be reused with a new dedicated name.
This is supported by adding a `register = "..."` argument once to a call to
`qualitative_hcl()`, `sequential_hcl()`, or `diverging_hcl()`:

@@

And can be used subsequently in `qualitative_hcl()` as well as the
qualitative \pkg{ggplot2} color scales (see Section\ \ref{sec:ggplot2}), e.g.,

@@

* The number of colors in the palette that was used during registration is not
  actually stored and can be modified subsequently. The same holds for arguments
  `alpha` and `rev`.
* When registering a new palette with an old name that was already available
  previously, the old palette gets overwritten. We recommend not to overwrite
  the palettes that are predefined in the package (albeit it is technically possible).
* The registration of a palette is only stored for the current session. When
  \proglang{R} is restarted and/or the \pkg{colorspace} package reloaded, only the predefined
  palettes from the package are available. Thus, to make a palette permanently
  available a registration \proglang{R} code like
  `colorspace::qualitative_hcl(3, palette = "set2", l = 80, register = "myset")`
  can be placed in your `.Rprofile` or similar startup scripts.

@@

```{r divergingx-palettes, fig.width = 8.0, fig.height = 4.25, out.width = "0.8\\textwidth", fig.cap = "Prespecified flexible diverging HCL palettes available in \\code{divergingx\\_hcl()} in \\pkg{colorspace}."}
divergingx_palettes(n = 7, plot = TRUE, nrow = 9)
```

@@

* \pkg{ColorBrewer.org} [@color:Harrower+Brewer:2003] as provided by \proglang{R} package \pkg{RColorBrewer} [@color:RColorBrewer].
  See `demo("brewer", package = "colorspace")`.
* \pkg{CARTO} colors [@color:CARTO] as provided by \proglang{R} package \pkg{rcartocolor} [@color:rcartocolor].
  See `demo("carto", package = "colorspace")`.
* The viridis palettes of @color:Smith+VanDerWalt:2015 developed for \pkg{matplotlib},
  as provided by \proglang{R} package \pkg{viridis} [@color:viridis].
  See `demo("viridis", package = "colorspace")`.
* The scientific color maps of @color:Crameri:2018 as provided by \proglang{R} package \pkg{scico} [@color:scico].
  See `demo("scico", package = "colorspace")`.

The graphics resulting from the demos can also be viewed online at
<http://colorspace.R-Forge.R-project.org/articles/approximations.html>.

In Figure\ \ref{fig:brewer-carto-viridis} `specplot()` panels (see also
Section\ \ref{sec:specplot}) are shown for two blue/green/yellow palettes (`RColorBrewer::brewer.pal(7, "YlGnBu")`
and `viridis::viridis(7)`) and two purple/red/yellow (`rcartocolor::carto_pal(7, "ag_Sunset")` and `viridis::plasma(7)`).
Each panel compares the hue, chroma, and luminance trajectories
of the original palettes (top swatch, solid lines) and their HCL-based approximations (bottom swatch, dashed lines).
The palettes are not identical but very close for most colors. Note that also the chroma trajectories from the HCL
palettes (green dashed lines) have some kinks which are due to fixing HCL coordinates at the boundaries of
admissible RGB colors.

Furthermore, Figure\ \ref{fig:brewer-carto-viridis} illustrates what sets the viridis palettes apart from
other sequential palettes. While the hue and luminance trajectories of `"Viridis"` and `"YlGnBu"` are very similar,
the chroma trajectories differ: While lighter colors (with high luminance) have low chroma for `"YlGnBu"`, they have
increasing chroma for `"Viridis"`. Similarly, `"ag_Sunset"` and `"Plasma"` have similar hue and luminance trajectories
but different chroma trajectories. The result is that the viridis palettes have rather high chroma throughout which
does not work so well for sequential palettes on a white/light background as all shaded areas convey high "intensity".
However, they work better on a dark/black background (see Figure\ \ref{fig:demoplot-dark}). Also, they might be a reasonable
alternative for qualitative palettes when grayscale printing should also work.

```{r brewer-carto-viridis, echo = FALSE, fig.width = 6, fig.height = 6, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of four palettes taken from \\pkg{ColorBrewer.org}, \\pkg{CARTO}, and \\pkg{viridis} (top swatch, solid lines) along with their HCL-based approximations (bottom swatch, dashed lines)."}
specplot(
  rev(RColorBrewer::brewer.pal(7, "YlGnBu")),
  sequential_hcl(7, "YlGnBu"),
  main = "ColorBrewer.org: YlGnBu"
)
specplot(
  viridis::viridis(7),
  sequential_hcl(7, "Viridis"),
  main = "viridis: Viridis"
)
specplot(
  rcartocolor::carto_pal(7, "ag_Sunset"),
  sequential_hcl(7, "ag_Sunset"),
  main = "CARTO: ag_Sunset"
)
specplot(
  viridis::plasma(7),
  sequential_hcl(7, "Plasma"),
  main = "viridis: Plasma"
)
```

@@

## Color swatches {#sec:swatchplot}

The function `swatchplot()` is a convenience function for displaying
collections of palettes that can be specified as lists or matrices of
hex color codes. Essentially, it is just a call to the base graphics
`rect()` function but with heuristics for choosing default labels, margins,
spacings, borders, etc. These are selected to work well for `hcl_palettes()`
and might need further tweaking in future versions. Thus,
Figures\ \ref{fig:hcl-properties}--\ref{fig:hcl-palettes} as well as
Figures\ \ref{fig:hcl-palettes-principles}--\ref{fig:hcl-palettes-diverging}
all use `swatchplot()` internally. For a simple stand-alone illustration
consider: `swatchplot("Palette" = sequential_hcl(5))`.

As a somewhat more complex example, a `swatchplot()` is created with three matrices
of sequential color palettes of blues, purples, reds, and greens. For all palettes,
luminance increases monotonically to yield a proper sequential palette. However,
the hue and chroma handling is somewhat different to emphasize different parts of the
palette.

* Single-hue: In each palette the hue is fixed and chroma decreases monotonically
  (along with increasing luminance). This is typically sufficient to clearly bring
  out the extreme colors (dark/colorful vs.\ light gray).
* Single-hue (advanced): The hue is fixed (as above) but the chroma trajectory is
  triangular. Compared to the basic single-hue palette above this allows to better
  distinguish the colors in the middle and not only the extremes.
* Multi-hue (advanced): As in the advanced single-hue palette the chroma trajectory
  is triangular but additionally the hue varies slightly. This can further enhance
  the distinction of colors in the middle of the palette.

```{r swatch-brpg, fig.width = 11.5, fig.height = 3, fig.cap = "Variations of blue, purple, red, and green palettes with single hue and monotonic chroma (left), single hue and triangular chroma (center), and multiple hues and triangular chroma (right)."}
bprg <- c("Blues", "Purples", "Reds", "Greens")
swatchplot(
"Single-hue"		= t(sapply(paste(bprg, 2), sequential_hcl, n = 7)),
"Single-hue (advanced)" = t(sapply(paste(bprg, 3), sequential_hcl, n = 7)),
"Multi-hue (advanced)"  = t(sapply(bprg,	   sequential_hcl, n = 7)),
nrow = 5, line = 5)
```

## HCL (and RGB) spectrum {#sec:specplot}

@@

The function first transforms a given color palette to its HCL (`polarLUV()`)
coordinates. As the hues for low-chroma colors are not (or only poorly) identified,
by default a smoothing is applied to the hues. Also, to avoid jumps from 0 to 360 or vice
versa, the hue coordinates are shifted suitably.
By default, the resulting trajectories in the HCL spectrum are visualized by a simple line plot:

@@

As spectrum plots have already been used for illustration in Figures\ \ref{fig:allplots-qualitative}
(for a qualitative palette) as well as \ref{fig:allplots-sequential} and \ref{brewer-carto-viridis}
(for sequential palettes), this section only provides a couple of additional illustrations.
The diverging `"Green-Brown"` palette is depicted in the left panel of Figure\ \ref{fig:specplot}.
It simply combines a green and a brown/yellow sequential single-hue palette, both with triangular chroma trajectory.
Hue is constant in each "arm" of the palette and the chroma/luminance trajectories are rather balanced between both arms.
In the center the palette passes through a light gray (with zero chroma) as the neutral value.
By including the corresponding RGB spectrum in the top panel, it also becomes apparent that
choosing such well-balanced palettes through trajectories in RGB color space is not straightforward.
This balanced palette -- based on relatively simple HCL trajectories -- is contrasted
with a poorly-balanced palette -- based on simple linear RGB trajectories in the right panel of
Figure\ \ref{fig:specplot}. This depicts the RGB and HCL spectrum of the 
(in)famous RGB rainbow palette. [See @color:Hawkins+McNeall+Stephenson:2014 for a
plea why the RGB rainbow palette should be avoided in almost all scientific graphics.]

```{r specplot, fig.width = 5, fig.height = 7, out.width = "0.49\\textwidth", fig.cap = "HCL spectrum of the balanced diverging \\code{\"Green-Brown\"} palette (left panel) and the (in)famous and rather unbalanced \\code{rainbow()} palette (right panel)."}
specplot(diverging_hcl(100, "Green-Brown"), rgb = TRUE)
specplot(rainbow(100), rgb = TRUE)
```

The RGB spectrum of the rainbow palette shows that the trajectories are quite simple in RGB space but
lead to substantial variations in chroma and (more importantly) luminance.
This is why this palette is not suitable for encoding underlying data in statistical graphics.
See also the related discussion of color vision deficiency in Section\ \ref{sec:color_vision_deficiency}.

## Trajectories in HCL space {#sec:hclplot}

@@

Therefore, the `hclplot()` is another visualization of the HCL coordinates associated
with a palette. It does so by collapsing over one of the coordinates (either the hue $H$
or the luminance $L$) and displays a heatmap of colors combining the remaining
two dimensions. The coordinates for the given color palette are highlighted
to bring out its trajectory. In case the hue is really fixed (as in single-hue
sequential palettes) or the luminance is really fixed (as in the qualitative
palettes), collapsing is straightforward. However, when the coordinate that is
collapsed over is actually not constant in the palette, a simple bivariate linear
model is used to capture how the collapsed coordinate varies along with the
two displayed coordinates.

@@

* `type = "qualitative"` if luminance is approximately constant.
* `type = "sequential"` if luminance is monotonic.
* `type = "diverging"` if luminance is diverging with two monotonic "arms" in the trajectory.

Thus, for qualitative palettes -- where luminance and chroma are fixed -- the varying
hue is displayed in a projection onto the hue-chroma plane at a given fixed luminance
(Figure\ \ref{fig:hcl-qualitative}):

```{r hcl-qualitative, fig.width = 4, fig.height = 4, dev = "png", out.width = "0.5\\textwidth", fig.cap = "Hue-chroma plane with luminance fixed at $L = 70$ along with a qualitative palette with varying hue $H$ and chroma fixed at $C = 50$."}
hclplot(qualitative_hcl(9, c = 50, l = 70))
```

@@

```{r hcl-sequential, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with variations of blue sequential single-hue palettes. Left: Linear chroma for $H = 260$. Center: Triangular chroma for $H = 245$. Right: Power-transformed triangular chroma for $H = 245$."}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = 260, c = 80, l = c(35, 95), power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95),
  power = 1))
hclplot(sequential_hcl(7, h = 245, c = c(40, 75, 0), l = c(30, 95),
  power = c(0.8, 1.4)))
```

Note that for $H = 260$ it is possible to go to dark colors (= low luminance) with
high chroma while this is not possible to the same extent for $H = 245$. Hence,
chroma has to be decreased when proceeding to the dark low-luminance colors.

@@

```{r hcl-multi, fig.width = 11.5, fig.height = 4.0, dev = "png", fig.cap = "Luminance-chroma plane with blue multi-hue palette and triangular chroma (left), blue-yellow multi-hue palette and linear chroma (center), and diverging blue-red palette with balanced linear chroma."}
par(mfrow = c(1, 3))
hclplot(sequential_hcl(7, h = c(260, 220), c = c(50, 75, 0),
  l = c(30, 95), power = 1))
hclplot(sequential_hcl(7, h = c(260, 60), c = 60, l = c(40, 95),
  power = 1))
hclplot(diverging_hcl(7, h = c(260, 0), c = 80, l = c(35, 95),
  power = 1))
```

@@

## Demonstration of statistical graphics {#sec:demoplot}

```{r demoplot-all, echo = FALSE, fig.width = 8.5, fig.height = 8.5, fig.cap = "All built-in \\code{demoplot} types with the same \\code{sequential\\_hcl(5, \"Heat\")} palette."}
par(mfrow = c(3, 3))
cl <- sequential_hcl(5, "Heat")
for (i in c("map", "heatmap", "scatter", "spine", "bar", "pie", "perspective", "mosaic", "lines")) {
  demoplot(cl, type = i)
  mtext(side = 3, text = i, line = 1.5)
}
```

To demonstrate how different kinds of color palettes work in different
kinds of statistical displays, `demoplot()` provides a simple convenience
interface to some base graphics with (mostly artificial) data sets.
As a first overview, Figure\ \ref{fig:demoplot-all} displays all built-in
demos with the same sequential heat colors palette: `sequential_hcl(5, "Heat")`.
All types of demos can, in principle, deal with arbitrarily many colors from any palette
but clearly the graphics differ in various respects such as:

@@

```{r demoplot-qualitative, fig.width = 8.5, fig.height = 3.0, fig.cap = "Examples for \\code{demoplot()} with different \\code{qualitative\\_hcl()} palettes."}
par(mfrow = c(1, 3))
demoplot(qualitative_hcl(4, "Pastel 1"), type = "pie")
demoplot(qualitative_hcl(4, "Set 2"),    type = "scatter")
demoplot(qualitative_hcl(4, "Dark 3"),   type = "lines")
```

@@

```{r demoplot-sequential, fig.width = 8.5, fig.height = 3.0, fig.cap = "Examples for \\code{demoplot()} with different \\code{sequential\\_hcl()} palettes."}
par(mfrow = c(1, 3))
demoplot(sequential_hcl(99, "Purple-Blue"), type = "heatmap")
demoplot(sequential_hcl(99, "Reds"),        type = "perspective")
demoplot(sequential_hcl( 4, "Viridis"),     type = "spine")
```

@@

```{r demoplot-diverging, fig.width = 8.5, fig.height = 3.0, fig.cap = "Examples for \\code{demoplot()} with different \\code{diverging\\_hcl()} palettes."}
par(mfrow = c(1, 3))
demoplot(diverging_hcl(99, "Tropic", power = 2.5), type = "map")
demoplot(diverging_hcl( 5, "Green-Orange"),        type = "mosaic")
demoplot(diverging_hcl( 5, "Blue-Red 2"),          type = "bar")
```

@@

```{r demoplot-dark, fig.width = 8.5, fig.height = 6.0, fig.cap = "Examples for \\code{demoplot()} with different palettes that work well on a black/dark background."}
par(mfrow = c(2, 3), bg = "black")
demoplot(sequential_hcl(9, "Oslo"), "heatmap")
demoplot(sequential_hcl(9, "Turku"), "heatmap")
demoplot(sequential_hcl(9, "Inferno", rev = TRUE), "heatmap")
demoplot(qualitative_hcl(9, "Set 2"), "lines")
demoplot(diverging_hcl(9, "Berlin"), "scatter")
demoplot(diverging_hcl(9, "Cyan-Magenta", l2 = 20), "lines")
```

@@

Using the physiologically-based model for simulating color vision deficiency (CVD)
of @color:Machado+Oliveira+Fernandes:2009 different kinds of limitations can be
emulated: deuteranope (green cone cells defective), protanope (red cone cells defective),
and tritanope (blue cone cells defective).

Below we briefly describe our \proglang{R} interface to these emulation techniques
and show them in practice for a heatmap with sequential palette. Another example
with diverging palette is available at
<http://colorspace.R-Forge.R-project.org/articles/color_vision_deficiency.html>.
Finally, CVD emulation is particularly useful for bringing out why the RGB rainbow
palette is almost always a bad choice in scientific displays. See
<http://colorspace.R-Forge.R-project.org/articles/endrainbow.html> for further
illustrations.

## R functions
